A generalization of the Bradley-Terry model for draws in chess

### 2020-10-05 inst/jebo_chess_rebuttal.txt

An excellent observation.  In the chess world the pre-eminent strength
measure would be the Elo rating in which the winning player takes
points from the loser based on the pre-match Elo rating difference.
Elo exhibits an emergent phenomenon in which sudden changes of ability
are not reflected instantly but have an exponentially damped response.

@@

This interpretation of Moul and Nye is very close to my own, and
re-reading my work I realise that the sentence in question was poorly
phrased and indeed clearly (but unintentionally) misleading.  I have
rephrased the sentence in question, and added a few more points of
reference to Moul and Nye, in BLUE.

> 4) The paper concludes that the estimates derived from the 1962
> Interzonals and the 1963 Candidates strongly suggest that the "draw
> monster" won games involving two Soviet players more frequently than
> games that involved at least one non-Soviet.  This is consistent with
> a Soviet cartel, but it could also be the result of Soviet-specific
> styles of play.  Moul and Nye (2009) test the Soviet style-of-play
> hypothesis by considering international tournaments when the format
> changed from all-play-all to knockout and all Soviet tournaments.
> Both comparisons yield results consistent with a Soviet draw-cartel in
> international all-play-all tournaments and rejecting the Soviet
> style-of-play hypothesis.  By only considering tournaments that were
> international and all-play-all, the author forgoes this critical
> source of identification.
> 
> In the cartel example, the author also neglects the advantage that
> presumably accrued to Soviet players when they played non-Soviet
> players in all-play-all international tournaments.  This would suggest
> another "monster" to be estimated, one that benefits the Soviet player
> when playing a non-Soviet opponent.  Without such an addition, Soviet
> strengths will be overstated.
>

(comments addressing this comment in DARK GREEN) This is an excellent
suggestion, one which I feel nicely rounds out the analysis.  I have
included a "rest monster" whose strength aids players following a
collusive draw.  Reified Bradley-Terry allows one not only to
establish the existence of such a rest monster, but also to quantify
its strength and I have done just this in the revision, and compared
the findings with a classical McNemar's test.

### 2020-10-08 inst/jebo_chess_revision2.tex
correction of pvalues and rephrasing

\begin{abstract}
\definecolor{DarkGreen}{RGB}{14,159,4}
In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing a second entity whose strength is added
to that of the white player.  These techniques afford insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  The likelihood
functions arising are easily optimised numerically.  I analyse a
number of datasets of chess results, including the infamous 1963 World
Chess Championships, in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work.  \textcolor{DarkGreen}{I also present
  scoreline evidence for the effectiveness of such a drawing cartel:
  noncollusive games are less detrimental to future play than
  collusive games ($p<10^{-6}$).}
\end{abstract}

@@

\textcolor{DarkGreen}{The Bradley-Terry model together with a white
  player advantage, draw monster, and rest monster, estimates the
  strength of the rest monster at about 26\% (compare the draw monster
  at 24\% and the White advantage at about 3\%).  The null of the rest
  monster having zero strength is rejected with a likelihood ratio of
  about $e^{11.9}$ or a $p$-value of about $10^{-6}$.  It is an
  interesting observation that neither Table~\ref{restmonsterfisher},
  nor reified Bradley-Terry, admit a White advantage (one-sided
  McNemar test, $p=\frac{1}{8}$, likelihood ratio $1.609$
  respectively).  }

@@

\textcolor{DarkGreen}{The ``rest monster'' concept is due to a
  suggestion by an insightful JEBO referee}

>>>>

\begin{abstract}
\definecolor{DarkGreen}{RGB}{14,159,4}
In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing a second entity whose strength is added
to that of the white player.  These techniques afford insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  The likelihood
functions arising are easily optimised numerically.  I analyse a
number of datasets of chess results, including the infamous 1963 World
Chess Championships, in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work.  \textcolor{DarkGreen}{I also present
  scoreline evidence for the effectiveness of such a drawing cartel:
  noncollusive games are less detrimental to future play than
  collusive games ($p<10^{-5}$).}
\end{abstract}

@@

\textcolor{DarkGreen}{The Bradley-Terry model together with a white
  player advantage, draw monster, and rest monster, estimates the
  strength of the rest monster at about 26\% (compare the draw monster
  at 24\% and the White advantage at about 3\%).  The null of the rest
  monster having zero strength is rejected with a likelihood ratio of
  about $e^{11.9}$ or a $p$-value of about $7\times 10^{-6}$.  It is
  an interesting observation that neither
  Table~\ref{restmonsterfisher}, nor reified Bradley-Terry, admit a
  White advantage (one-sided McNemar test, $p=\frac{1}{8}$, likelihood
  ratio $1.609$ respectively).  }

@@

\textcolor{DarkGreen}{The ``rest monster'' concept is due to a
  suggestion by an insightful JEBO referee.}
